---
title: "comparing two data frames: one #rstats, many ways!"
author: Sharla Gelfand
date: "2020-02-17"
slug: comparing-two-dfs
---

today i tweeted asking about packages to compare two data frames (specifically, their variable names and types):

```{r echo=FALSE}
blogdown::shortcode("tweet", "1229512910591885317")
```

judging by the ratio (18 replies to 38 likes!) at the time of writing, *yes* there are definitely packages to do this! in this post, i want to explore the (many) different and excellent methods for comparing two data frames in #rstats.

# the data

first, what are we talking about? imagine i have some data coming in, and i know what it should look like. it should look like `iris`! `r emo::ji("flower")`

```{r, include = FALSE}
iris <- tibble::as_tibble(iris)
options(tibble.print_max = 5, tibble.print_min = 5)
knitr::opts_chunk$set(message = FALSE)
```

```{r}
iris
```

so, `iris` has 5 columns: Sepal.Length, Sepal.Width, Petal.Length, and Species. the column types are: numeric/double (times 4!) and factor.

the data i'm bringing in could have a number of problems.

it could be missing a column:

```{r, message = FALSE}
library(dplyr)

iris_missing_cols <- iris %>%
  select(-Sepal.Length)

iris_missing_cols
```

it could have an extra column:

```{r}
iris_extra_cols <- iris %>%
  mutate(Specieso = Species)

iris_extra_cols
```

it could be missing a column *and* have an extra column:

```{r}
iris_missing_extra_cols <- iris %>%
  select(-Sepal.Length) %>%
  mutate(Specieso = Species)

iris_missing_extra_cols
```

it could have a column of the wrong class:

```{r}
iris_wrong_class <- iris %>%
  mutate(Species = as.character(Species))

iris_wrong_class
```

or it could have all three!

```{r}
iris_missing_extra_cols_wrong_class <- iris_missing_extra_cols %>%
  mutate(Species = as.character(Species))

iris_missing_extra_cols_wrong_class
```

one last thing is not a problem, but i'm curious to see if the packages i explore will think it is. this is when the column *order* is different:

```{r}
iris_diff_order <- iris %>%
  select(Species, everything())

iris_diff_order
```

# what i dream of `r emo::ji("cloud")`

in the tweet i mentioned i'm looking for packages that will provide me with nice, descriptive errors (or robust data to create my own) describing what's different between the data sets.

what does this look like? two main things.

1. saying what columns are missing (and what's optionally, what's extra).

    for `iris` vs `iris_missing_cols`, i want to know that `iris_missing_cols` is missing a column called Sepal.Length. 
    
    for `iris` vs `iris_extra_cols`, i want to know that `iris_extra_cols` has an extra column called Specieso.
    
    for `iris` vs `iris_missing_extra_cols`, i want to know that `iris_missing_extra_cols` is missing Sepal.Length and has an extra column Specieso.
    
2. saying what classes are different.

    for `iris` vs `iris_wrong_class`, i want to know that Species *should be* a factor, but it's a character.

ok, putting them together, three things.

3. saying what columns are missing, what's extra, and what classes are different.

    for `iris` vs `iris_missing_extra_cols_wrong_class`, i want to know that `iris_missing_extra_cols_wrong_class` is missing Sepal.Length, has an extra column Specieso, and that Species should  be a factor but instead is a character.

finally, if the data frames are the same, it would be nice to have a function that returns TRUE or just the original data.

so! what can i use to get this?

# the functions

## `dplyr::all_equal()`

let's check out some comparisons using dplyr::all_equal(), suggested by [Dilsher Dhillon](https://twitter.com/dhillon_stats/status/1229517066534817793)!

first, what happens if the data frames are the same?

```{r}
all_equal(iris, iris)
```

nice

if they are the same, but a different order?

```{r}
all_equal(iris, iris_diff_order)
```

wena

if there are missing and extra columns?

```{r}
all_equal(iris, iris_missing_extra_cols)
```

if there's an incorrect variable class?

```{r}
all_equal(iris, iris_wrong_class)
```

what about wrong columns and the wrong class?

```{r}
all_equal(iris, iris_missing_extra_cols_wrong_class)
```

looks like however `dplyr::all_equal()` works, reporting on the missing/extra columns is prioritized over reporting on the incorrect variable class. this is not great but not a *huge* deal to me - the function could report on the wrong columns, they get fixed, and *then* it reports on the wrong class.

overall, i do like the nice printing, e.g. "cols in y but not in x", and these could probably be str_replace-ified in a function to supply the actual data frame names (e.g. iris vs iris_wrong_class).

i'm running the version of dplyr on CRAN - i had a dev version installed earlier and some of the messages [looked different](https://github.com/tidyverse/dplyr/blob/master/R/all-equal.r#L74). if these will be changing (that's ok!) i wouldn't want to rely on altering the string output for creating my own errors, since the strings could chang, and i don't necessarily want to require dplyr 1.0.0 in a package just for all_equal strings :~)

## `janitor::compare_df_cols()`

i consider myself a big fan of janitor but there's always new functions in there that surprise me! thank you to [Rebecca Janis](https://twitter.com/rbjanis5/status/1229523325564727298) for mentioning this (and sharing her own [function for comparing column classes](https://gist.github.com/rbjanis/44dbcecb2c96c11defdb7892ad8158fb))!

let's go in the same order. what happens if the data frames are the same?

```{r, message = FALSE}
library(janitor)

compare_df_cols(iris, iris)
```

it just returns a comparison of the columns ()
(what's in both data frames, and their classes in each).

you can set an option only to return things that *don't* match (or things that do):

```{r}
compare_df_cols(iris, iris, return = "mismatch")
```

(in this case, nothing).

i assume that this will be robust to different order of columns, but let's double check:

```{r}
compare_df_cols(iris, iris_diff_order)
```

cool

what about missing or extra columns?

```{r}
compare_df_cols(iris, iris_missing_extra_cols)
```

i like this because it shows all columns in both, and what their types are, by data frame (NA if the column isn't in that df)

strangly enough, a "mismatch" here is a mismatch in class, and NA does not count as a mismatch:

```{r}
compare_df_cols(iris, iris_missing_extra_cols, return = "mismatch")
```

and does count as a match `r emo::ji("upside_down_face")`

```{r}
compare_df_cols(iris, iris_missing_extra_cols, return = "match")
```

what about if the variable classes don't match?

```{r}
compare_df_cols(iris, iris_wrong_class, return = "mismatch")
```

i like this because you can easily construst your own message from this structure, using glue:

```{r, message = FALSE}
library(glue)

compare_df_cols(iris, iris_wrong_class, return = "mismatch") %>%
  mutate(message = glue("Column `{column_name}`: expected class {iris}, got class {iris_wrong_class}")) %>%
  pull(message)
```

and the resulting data frame can be filtered manually when the filters from "mismatch" aren't what i want, to see all differences:

```{r}
compare_df_cols(iris, iris_missing_extra_cols_wrong_class) %>%
  filter(is.na(iris) | is.na(iris_missing_extra_cols_wrong_class) |
    iris != iris_missing_extra_cols_wrong_class)
```

## `vetr::alike()`

next up is `vetr::alike()`! this is a new package to me, excited to see what it can do `r emo::ji("mag")`! thanks to [Brodie G](https://twitter.com/BrodieGaslam/status/1229520931451789312) for suggesting (and creating!) it.

again, let's start off checking on data frames that match:

```{r}
library(vetr)

alike(iris, iris)
```

cool, what about different order of columns?

```{r}
alike(iris, iris_diff_order)
```

uh oh

what about a data frame with missing and extra columns?

```{r}
alike(iris, iris_missing_extra_cols)
```

this only tells me that one of the column names is wrong (the first one), and not what's extra.

it *can* identify when there's extra columns:

```{r}
alike(iris, iris_extra_cols)
```

but doesn't say what they are. similarly to `dplyr::all_equal()`, it seems like it prioritizes one kind of reporting (an incorrect column name) and doesn't indicate other issues.

let's look at an incorrect class:

```{r}
alike(iris, iris_wrong_class)
```

i like this format - says what it should be, and what it actually is.

## `testthat::expect_equal()`

this one is my idea `r emo::ji("innocent")` because i love testthat

when the data frames are the same, `expect_equal()` returns:

```{r}
library(testthat)

expect_equal(iris, iris)
```

nothing. well actually, it returns the data frame invisibly:

```{r}
x <- expect_equal(iris, iris)

x
```

but doesn't print it automatially, probably since it's usually used within a `test_that()` call.

when there are missing/extra columns:

```{r, error = TRUE}
expect_equal(iris, iris_missing_extra_cols)
```

the reporting is very similar (the same?) as from `dplyr::all_equal()` `r emo::ji("thinking")` are these the same functions?

```{r, error = TRUE}
expect_equal(iris, iris_wrong_class)
```

ok yes, the same as `dplyr::all_equal()` but returning an error instead of a string. it probably just calls `all_equal()`. maybe i shouldn't have bragged that this one was *my* idea `r emo::ji("joy")`

## `diffdf::diffdf()`

[Marianna Foos](https://twitter.com/MariannaFoos/status/1229570688073322496) swept in while i was writing this post and suggested diffdf!

```{r}
library(diffdf)

diffdf(iris, iris)
```

thank you, diffdf!

```{r}
diffdf(iris, iris_diff_order)
```

i agree!

```{r}
diffdf(iris, iris_missing_extra_cols)
```

shit ok! we got DIFFERENCES !!

```{r}
diffdf(iris, iris_wrong_class)
```

i don't know what a mode is (something related to factors, i guess!) but i like that it reports the different classes in Species.

this function also returns a list of data frames with issues invisibly (i surpressed warnings because it was yelling ok)

```{r}
x <- diffdf(iris, iris_missing_extra_cols_wrong_class, suppress_warnings = TRUE)

x[["VarClassDiffs"]]
x[["ExtColsBase"]]
x[["ExtColsComp"]]
```

this is similar to the kind of information you get from `janitor::compare_df_cols()`, and i like it!

# comparing the functions

i have a bunch of functions to compare, on a bunch of criteria!

1. iris == iris `r emo::ji("yum")`
2. iris in a different order is still iris
3. detects missing columns
4. detects extra columns
5. detects missing *and* extra columns
7. detects difference in class
8. detects different columns *and* difference in class
9. nice strings to use for messages
10. returns data (that i can use to BYOmessages)

i'm not going to include `testthat::test_equal()` in the comparison, since it's so similar (read: basically the same) to `dplyr::all_equal()`

```{r, echo = FALSE}
yes <- "\U2705"
no <- "\U274C"
tibble::tribble(
  ~metric, ~`dplyr::all_equal()`, ~`janitor::compare_df_cols()`, ~`vetr::alike()`, ~`diffdf::diffdf()`,
  "iris is iris", yes, yes, yes, yes,
  "column swapped iris is iris", yes, yes, no, yes,
  "missing columns", yes, yes, yes, yes,
  "extra columns", yes, yes, yes, yes,
  "missing *and* extra columns", yes, yes, no, yes,
  "difference in class", yes, yes, yes, yes,
  "different columns *and* classes", no, yes, no, yes,
  "nice strings to use for messages", yes, no, paste0(yes, no), no,
  "returns data on differences", no, yes, no, yes
) %>%
  knitr::kable(format = "html", align = "lccccc", col.names = c("", names(.)[-1])) %>%
  kableExtra::kable_styling(full_width = TRUE)
```

# what to do what to do

ok, so what did i end up doing? combining things, of course!

i loved learning about new packages and seeing what's available in packages i already know. i am inclined to use stuff from dplyr and janitor because they're often already dependencies in my work.

so, i decided to cobble something together using the data available from `janitor::compare_df_cols()`, and the basic check of `dplyr::all_equal()` as a jumping off point, and  i borrowed some of the message/error/etc wording from the packages i explored. i also have my function return the original data if all checks pass so that it can easily be used in a pipeline!

i won't belabour you with the [source code](https://gist.github.com/sharlagelfand/4d2713ca2cdd106bdecfca0130007a5f), but here's how it works:

```{r, include = FALSE}
sharla_diff <- function(df, expected_df) {
  data_as_expected <- dplyr::all_equal(expected_df, df)

  if (!isTRUE(data_as_expected)) {
    data_diffs <- janitor::compare_df_cols(expected_df, df)

    cols_mismatch <- dplyr::filter(data_diffs, is.na(expected_df) | is.na(df))

    extra_cols <- cols_mismatch %>%
      dplyr::filter(is.na(expected_df)) %>%
      dplyr::pull(column_name)

    missing_cols <- cols_mismatch %>%
      dplyr::filter(is.na(df)) %>%
      dplyr::pull(column_name)

    type_mismatch <- dplyr::filter(data_diffs, expected_df != df)

    if (length(extra_cols) > 0) {
      warning("`", deparse(substitute(df)), "`", " contains extra column(s): ",
        glue::glue_collapse(extra_cols, sep = ", "), ".",
        call. = FALSE
      )
    } 
    
    if (length(missing_cols) > 0) {
      stop("`", deparse(substitute(df)), "`", " is missing column(s): ", glue::glue_collapse(missing_cols, sep = ", "), ".",
        call. = FALSE
      )
    } 
    
    if (nrow(type_mismatch) > 0) {
      type_mismatch_errors <- type_mismatch %>%
        dplyr::mutate(error = glue::glue("{column_name} should be class {expected_df}, is {df}")) %>%
        dplyr::pull(error) %>%
        as.character() %>%
        paste0("\n")

      stop("`", deparse(substitute(df)), "`", " column types are not as expected. These columns have issues:\n", type_mismatch_errors,
        call. = FALSE
      )
    }
  }

  df
}
```


```{r, error = TRUE}
sharla_diff(iris, iris)

sharla_diff(iris, iris_diff_order)

sharla_diff(iris, iris_missing_extra_cols)

sharla_diff(iris, iris_wrong_class)

sharla_diff(iris, iris_missing_extra_cols_wrong_class)
```

`sharla_diff()` errors when it's missing columns, so it doesn't have the chance to look for the wrong class in `iris_missing_extra_cols_wrong_class` (like `dplyr::all_equal()`!). as i mentioned, that's not a huge deal for me.

i'm sure this function isn't perfect and will be changed as i see issues with it. that's it for now! please see the honourable mentions appendix if you think i snubbed your package ok!!!

# honourable mentions

## packages for comparing values

some people suggested packages for comparing the actual *values* within a data frame instead of the data frame name/classes. that's not what i was looking for, but helpful anyways! 

if you want to do that:

`visdat` in general is such an awesome package for visually understanding your data. [Nick Tierney](https://twitter.com/nj_tierney/status/1229531977100840961) is its creator (and suggestor!), specifically suggesting [visdat::vis_compare()](https://visdat.njtierney.com/#using-vis_compare) for visually comparing values.

[dataCompareR::rCompare()](https://cran.r-project.org/web/packages/dataCompareR/vignettes/dataCompareR.html) can be used for comparing values very robustly. thank you to [Robert Campbell](https://twitter.com/robertc_93/status/1229544225060118528) for the suggestion!

a little unrelated, but [diffobj](https://cran.r-project.org/web/packages/diffobj/vignettes/diffobj.html) looks very cool for visualizing diffs of objects (duh, hence the name) within your console. ty to [Tokhir Dadaev](https://twitter.com/zx8754/status/1229514956112113666) and [Brodie Gaslam](https://twitter.com/BrodieGaslam/status/1229518614853750784) for mentioning (and Brodie for creating - a double for both in this post!).

## daff::diff_data() and arsenal::comparedf()

ok, thank you and no offence to everyone ([Erin Grand](https://twitter.com/astroeringrand/status/1229520088073736193), [Antoine Fabri](https://twitter.com/antoine_fabri/status/1229527662202363911), [Eric Polley](https://twitter.com/ecpolley/status/1229557115812970498), [Benjamin Gowan](https://twitter.com/Benjaming_G/status/1229514790734663680)) that suggested these!!! [daff::diff_data()](https://github.com/edwindj/daff/blob/master/README.md) doesn't render in rmarkdown and [arsenal::comparedf()](https://cran.r-project.org/web/packages/arsenal/vignettes/comparedf.html) is a little too robust for what i'm looking for.

either way, both methods provide more information than i want and don't really lend themselves to easy creation of errors or warnings. if you are looking for a full report (tm) of what is different between two data frames, these might be for you!

## vctrs::vec_assert()

according to [Hadley](https://twitter.com/hadleywickham/status/1229517175062269952), `vctrs::vec_assert()` might one day create the kind of messages i want about classes (e.g. expected Date, got character). i look forward to it (and to learning vctrs someday)!

ok that's *really* it! bye!
